\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-01-18 response.tex

		Page 23. Replace 'oberve' by 'observe'.

		\textcolor{blue}{Done}

		Page 24. Petr Hliněný (together with his student) has announced that
		the answer to the open question is yes, with $C=3$ and $\ell=O(k^2)$.

\subsection*{Referee \#1}

This paper proves that if a class of graphs F admits a product structure
theorem, then the class of all graphs which have a 1-subdivision in F admits
a product structure theorem as well. The main theorem, Theorem 3, while
equivalent to the above statement using standard techniques, is stated in
a different way which is considerably more useful for applications and for
obtaining good bounds. As applications of Theorem 3, the paper shows that
the following classes admit a product structure theorem:

\begin{itemize}
\item graphs drawn on a fixed surface with each edge in $\leq k$ crossings,
\item map graphs of degree $\leq k$ graphs embedded in a fixed surface,
\item string graphs with each curve in $\leq k$ intersections,
\item $k$-nearest neighbour graphs in the plane, and
\item powers of classes of graphs of degree $\leq k$ which admit a product structure theorem.
\end{itemize}

All but the last bullet point are shown to follow from the first.
The main results in the paper are correct, strong, and of high interest,
and the paper should definitely be accepted to JCTB after some
(fairly significant) revisions. In particular, the proof of Theorem 3 is
difficult to read due to the notation, and the paper could be substantially
streamlined without losing the above results. Detailed suggestions follow.

Major comments:

1. Lemma 3: This lemma is really about “normalised” tree decompositions, and says that for each $x \in V(T)$, the set $V(T_x)$ has at most $t$ neighbours in $H$. Call this property (T3) and remove (Y1)  through (Y5), which should now be obvious.

2. Proof of Theorem 9: There is too much notation throughout. As
one example, it is hard to remember which of $T_x$, $Y_x$, $V_x$, $F_x$, $N_x$,
$S_x$, $X_v$, $B_x$, $C_x$ is which. If $\phi(S)$ is written for the subset of $V(G)$
corresponding to a set $S \subseteq V(H)$ and appropriate shorthand and
standard notation for neighbourhoods are used, then $Y_x$, $V_x$, $F_x$, and
$N_x$ can be replaced by $\phi(x)$, $\phi(T_x)$, $N(\phi(T_x))$, and $V(G) \setminus N[\phi(T_x)]$.
The set $X_v$ is only used in the statement of Claim~1, so it does not
need a name; say what these vertices are explicitly. It may be more
indicative to write $\phi'(x)$ for $S_x$ and $B'_x$ for $C_x$.

3. Theorem 10: Unfortunately I have to suggest removing it. I don’t see
the motivation for making H planar or for improving the constants.
Are the improved constants at all close to tight? Is there another
reason for including it?

4. Sections 4 and 5: I think as they are currently organized they distract from the purpose of the paper, as stated in the introduction, to
“prove product structure theorems for several non-minor-closed classes
of interest.” Which of Corollaries 1–4 and Theorems 12–17 are most
important? I would remove any mention of the applications from the
examples section, put that section first, and state in that section a single main theorem with all of the best bounds on the product structure.
The applications section is mostly a survey and I think should be
focused on the most exciting new corollaries. Theorems 18 and 19
should be in the applications section.

Other comments:

1. abstract: “This leads to analogous results for map graphs, string graphs, graph powers, and nearest neighbour graphs.” $\longrightarrow$ Add “certain types of” or be more specific. Same comment for immediately after Theorem 4.

2. page 1: “A tree-decomposition T consists of a tree T and a collection
T ” $\longrightarrow$ Delete the first T and define a T-decomposition.

3. pages 2 – 4: The paragraph following Theorem 2 and Section 1.3
(except for the fact that the apex assumption is necessary) is repetition
of things discussed earlier on in the introduction.

4. page 6, line 14: “Thus, it suffices to construct width-t tree decomposition that satisfies (T2).” $\longrightarrow$ should be (T1)

5. page 6, line 16: “Select any node $x \in V (T_0)$” $\longrightarrow$ “Select any nonempty node $x \in V (T_0)$”

6. page 6, last paragraph of Lemma 2: This isn’t quite right since you
have to be careful with x0.

7. page 8, item (i): “S has small layered width with respect to the layering
L” $\longrightarrow$ “S has small layered width with respect to the layering L of
G”. Perhaps mention something about how it is easy to find the new
layering.

8. general comment: Claims 1, 2, and 4 are especially belaboured by
notation as they are quite straightforward.

9. page 8, line 10: “We say that a vertex $w \in Y_x$ contributes a vertex
$v \in S_x$ if $v$ participates in $x$.” This needs to be fixed.

10. page 10, Claim 5: Indices could mostly be removed by continuing
to talk about T-ancestors instead. Move the definition of H+ and its
directed version up, along with the explanation of what will be proven,
so that it isn’t necessary to define the si
.
11. page 10, Claim 5: The name Pww0 is strange since in case 2. there is
no w
0 defined.

12. page 11, first paragraph of Section 3: “This section applies our main
results for shortcut systems to prove graph product structure theorem...” $\longrightarrow$ “to prove a graph product structure theorem...”

### 2022-01-18 response.tex

\documentclass[12pt]{article}
\listfiles
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,color}
\usepackage[margin=20mm]{geometry}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{paralist}
\usepackage{stmaryrd}

\setlength{\parindent}{0ex}
\setlength{\parskip}{1ex}

\begin{document}

\subsection*{Graph Product Structure for Non-Minor-Closed Classes\\
Vida Dujmovi\'c, Pat Morin, David R. Wood\\
Journal of Combinatorial Theory Series B (JCTB9758)\\
Response to Referee Reports}

@@

9. page 8, line 10: “We say that a vertex $w \in Y_x$ contributes a vertex
$v \in S_x$ if $v$ participates in $x$.” This needs to be fixed.

10. page 10, Claim 5: Indices could mostly be removed by continuing
to talk about T-ancestors instead. Move the definition of H+ and its
directed version up, along with the explanation of what will be proven,
so that it isn’t necessary to define the si
.
11. page 10, Claim 5: The name Pww0 is strange since in case 2. there is
no w 0 defined.

12. page 11, first paragraph of Section 3: “This section applies our main results for shortcut systems to prove graph product structure theorem...” $\longrightarrow$ “to prove a graph product structure theorem...”

\textcolor{blue}{done}

\subsection*{Referee \#2}

Page 3.  The sentence 'For each $vw \in E(G')$ there is a path $P$ in $G$
between v and we of length at most $k+1$'' is correct but should be made
more precise.  For example, there could be a short path between $v$ and
$w$ which does not involve any new dummy vertices.  You mean to say that
the 'subdivided path' between $v$ and $w$ has length at most $k+1$.

Page 3.  You should add the assumption that no three edges cross at
the same point (which you can assume by perturbing the embedding).
Otherwise, you will not get a $(k+1, 2)$-shortcut system.

Page 3. $P$ is used twice in the statement of Theorem 3.  It is probably
safer (and necessary?) to use a different variable for the second
occurrence.

Page 6. Regarding (T1), technically a subtree of a rooted tree can be
rooted at any vertex, so it is a bit imprecise to say that $T[x]$ is
rooted at $x$.  I suggest adding that a subtree of a tree rooted at $r$,
is always considered to be rooted at the vertex closest to $r$.

Page 11. Replace 'belong to $F_\delta'$ by 'belongs to $F_\delta$'.

Page 11. In footnote 11, what the authors call a 'closed curve' I
would call just a 'curve'.  To me, a closed curve satisfies the
additional condition that $f(0)=f(1)$.

\textcolor{blue}{done}

\textcolor{blue}{done}

\textcolor{blue}{We look forward to reading the paper.}

### 2022-01-18 k-planar.tex
added responses

This section applies our main results for shortcut systems to prove graph product structure theorem for graphs drawn with a bounded number of crossings per edge. Then we show how to improve the bounds in this case.

@@

\begin{thm}
\label{StringPartition}
For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$, 
and $t:= \binom{ \delta+5}{4}-1$ if $g\geq 1$ and $t:= \binom{ \delta+4}{3}-1$ if $g=0$. 
Then  every $(g,\delta)$-string graph: 
\begin{compactitem}
	\item is a subgraph of $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
	\item has queue-number $\qn(G^k) \leq 3 \ell \, 2^t - \ceil{\tfrac{3}{2}\ell}$,
\end{compactitem}
\end{thm}

@@

% % NOTE THAT MANUAL CHANGES HAVE BEEN MADE BELOW.
%
\begin{thebibliography}{37}
	\providecommand{\natexlab}[1]{#1}
	\providecommand{\url}[1]{\texttt{#1}}
	\expandafter\ifx\csname urlstyle\endcsname\relax
	\providecommand{\doi}[1]{doi: #1}\else
	\providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

@@

	\bibitem[Dujmovi{\'c} et~al.(2019)Dujmovi{\'c}, Joret, Micek, Morin, Ueckerdt,
	and Wood]{dujmovic.joret.ea:planar}
	Vida Dujmovi{\'c}, Gwena\"{e}l Joret, Piotr Micek, Pat Morin, Torsten Ueckerdt,
	and David~R. Wood.
	\newblock Planar graphs have bounded queue-number.
	\newblock In David Zuckerman, editor, \emph{Proc. 60th {IEEE} Annual Symp.
		Foundations Comput. Sci.} (FOCS 2019), pages 862--875. IEEE, 2019.
	\newblock \doi{10.1109/FOCS.2019.00056}.
	\newblock To appear in \emph{J. ACM}, \arXiv{1904.04791}.

	\bibitem[Dujmovi\'c et~al.(2020{\natexlab{a}})Dujmovi\'c, Esperet, Joret,
	Gavoille, Micek, and Morin]{DEJGMM}
	Vida Dujmovi\'c, Louis Esperet, Gwena\"el Joret, Cyril Gavoille, Piotr Micek,
	and Pat Morin.
	\newblock Adjacency labelling for planar graphs (and beyond).
	\newblock 2020{\natexlab{a}}.
	\newblock \arXiv{2003.04280}.

@@

	\bibitem[{Dujmovi{\'c} et~al.(2020)Dujmovi{\'c}, Esperet, Morin, Walczak, and Wood}]{DEMWW}
	Vida Dujmovi{\'c}, Louis Esperet, Pat Morin, Bartosz Walczak, and
 David~R. Wood.
	\newblock Clustered 3-colouring graphs of bounded degree.
	\newblock 2020.
	\newblock \arXiv{2002.11721}.

>>>>

This section applies our main results for shortcut systems to prove a product structure theorem for graphs drawn with a bounded number of crossings per edge. Then we show how to improve the bounds in this case.

@@

\begin{thm}
\label{StringPartition}
For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$, 
and $t:= \binom{ \delta+5}{4}-1$ if $g\geq 1$ and $t:= \binom{ \delta+4}{3}-1$ if $g=0$. 
Then  every $(g,\delta)$-string graph: 
\begin{compactitem}
	\item is a subgraph of $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
	\item has queue-number $\qn(G^k) \leq 3 \ell \, 2^t - \ceil{\tfrac{3}{2}\ell}$.
\end{compactitem}
\end{thm}

@@

% NOTE THAT MANUAL CHANGES HAVE BEEN MADE BELOW.

\begin{thebibliography}{37}
	\providecommand{\natexlab}[1]{#1}
	\providecommand{\url}[1]{\texttt{#1}}
	\expandafter\ifx\csname urlstyle\endcsname\relax
	\providecommand{\doi}[1]{doi: #1}\else
	\providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

@@

	\bibitem[Dujmovi{\'c} et~al.(2019)Dujmovi{\'c}, Joret, Micek, Morin, Ueckerdt,
	and Wood]{dujmovic.joret.ea:planar}
	Vida Dujmovi{\'c}, Gwena\"{e}l Joret, Piotr Micek, Pat Morin, Torsten Ueckerdt,
	and David~R. Wood.
	\newblock Planar graphs have bounded queue-number.
	\newblock \emph{J. ACM} 67(4):22, 2020.
	\newblock \doi{10.1145/3385731}.

	\bibitem[Dujmovi\'c et~al.(2020{\natexlab{a}})Dujmovi\'c, Esperet, Joret,
	Gavoille, Micek, and Morin]{DEJGMM}
	Vida Dujmovi\'c, Louis Esperet, Gwena\"el Joret, Cyril Gavoille, Piotr Micek,
	and Pat Morin.
	\newblock Adjacency labelling for planar graphs (and beyond).
	\newblock \emph{J. ACM} 68(6):42, 2021. 
	\newblock \doi{10.1145/3477542}.

@@

	\bibitem[{Dujmovi{\'c} et~al.(2020)Dujmovi{\'c}, Esperet, Morin, Walczak, and Wood}]{DEMWW}
	Vida Dujmovi{\'c}, Louis Esperet, Pat Morin, Bartosz Walczak, and
 David~R. Wood.
	\newblock Clustered 3-colouring graphs of bounded degree.
\newblock \emph{Combin. Probab. Comput.} 31(1):123--135, 2022.
\newblock \doi{10.1017/S0963548321000213}.
